A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-04-03 analysis/paper/paper_latex.Rmd
update paper

Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory. Since 2014, at least `r sum(implementations$is_ti, na.rm=TRUE)` so-called trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare. As a result, a comprehensive assessment of the performance and robustness of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.

In this study, we evaluated a total of  `r sum(methods$evaluated)` methods in terms of  cellular ordering, trajectory topology, code quality and user friendliness. Our results indicate that slingshot is the most accurate method to date, but other leading methods (e.g. TSCAN, Monocle DDRTree, SCORPIUS and cellTree) may perform better in specific use-cases. We therefore developed a set of guidelines to aid users select the most appropriate method. While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with ever-more complex use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).

# Introduction
Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017]. One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014]. Such dynamic processes can be modelled computationally using using trajectory inference (TI), which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018]. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow for more complex trajectory topologies such as cyclic or disconnected graphs. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017]. 
Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "implementations")`). All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce. Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type of topologies they can detect. Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along this fixed topology. 
Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods to be applicable on more use cases. Methods that perform topology inference are still in the minority, though current trends suggest this will soon change. A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is to our knowledge the only TI method currently able to deal with disconnected graphs.

<!-- §3: Problem statement -->
Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness. Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation, which compares a large set of methods against a large set of datasets, of TI methods is still lacking. This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about what method would optimally solve their problem. Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.

@@

In this study, we performed a comprehensive evaluation of  `r sum(methods$evaluated)` TI methods (`r ref("fig", "evaluation_overview", "a")`). The inclusion criterion for TI methods was primarily based on their free availability, the presence of a programming interface (`r ref("table", "implementations")`). The evaluation comprised three core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard, (ii) a quality control of the provided software and documentation, and (iii) combining both evaluations to create a set of practical guidelines for users.

`r ref("fig", "evaluation_overview")` **An overview of several key aspects of the evaluation.** a) A schematic overview of our evaluation pipeline. b) In order to make any two trajectories comparable with one another, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to one another, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. We then defined a set of metrics for comparing the likeness of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets came from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). To generate synthetic datasets, we simulated a gene regulatory network using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated a single-cell profiling experiment by matching the distributions of the synthetic data with real reference datasets.
<!-- Overall evaluation results -->
Across all datasets and evaluation metrics, we found that slingshot, TSCAN and Monocle DDRTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`). When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returns the most accurate linear trajectories, reCAT optimally identifies cycles, and Monocle DDRTree most accurately identifies trees. This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, slingshot, TSCAN and Monocle DDRTree, obtaining a better topological similarity score, while other methods, including SCORPIUS, slingshot and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).

@@

`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r sum(methods$evaluated)` TI methods.** a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use. b) The methods are ordered according to the overall score achieved in the benchmark evaluation. Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced. c) Overall performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation. Also listed are the quality control scores aggregated according to practicality and the different categories.

@@

<!-- Errors-->
Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g. Monocle DDRTree, SLICE, Wishbone), meaning these methods could rank even higher if not for the failed executions. Methods that do not scale well with respect to the number of cells or genes exceeded the pre-defined time or memory limits for the largest datasets (reCAT, SCOUP, StemID). For a few methods, the time or memory limits were exceeded too often, making the benchmarking results uncomparable to those of other methods (SCIMITAR, Ouija, Pseudogp).

<!-- Evaluation robustness -->
```{r}
correlation_tenth <- read_rds(result_file("task_sample_correlations.rds", "11-evaluation_robustness")) %>%
filter(n_tasks == round(max(n_tasks)/10)) %>% 
          pull(cor) %>% 
          mean()
top_method_rank_match <- read_rds(result_file("task_sample_ranks_match.rds", "11-evaluation_robustness")) %>%
filter(method_id == "slngsht") %>% 
filter(n_tasks == round(max(n_tasks)/10)) %>%
pull(rank_match)
random_method_max_tasks_top <- read_rds(result_file("task_sample_max_tasks_top.rds", "11-evaluation_robustness")) %>% 
filter(method_id == "scoup") %>% 
pull(max_n_tasks_top)

```

We further investigated the robustness of our evaluation across datasets and dataset sources. Method performance was moderately to highly correlated (0.5-0.9 depending on the trajectory type) between real and synthetic datasets (`r ref("sfig", "robustness_comparison", "a")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard. Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still very robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`). To further illustrate this, we looked at the scores in a randomly selected 10% subset of all datasets. The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets. On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best. For example, for SCOUP, a method ranked in the lower half, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method. Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large set of datasets, such as those included in our evaluation, is necessary to robustly compare the overall performance of methods.

## Relationships between method characteristics to performance

<!-- Inverse relation between edge flip and correlation -->
In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("sfig", "trajtype_handle_comparison")`). These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type. However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as embeddr, SCORPIUS and phenopath, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories. To further investigate the effect of trajectory complexity on performance, we divided the methods in two groups: non-branching methods (restricted to linear and cyclic topologies) and branching methods. While there were no significant differences in performance on linear datasets, branching methods produced significantly a better topology, but a less accurate cellular ordering on datasets containing at least one branching point (`r ref("sfig", "linear_split_comparison")`). Together, this indicates that current non-linear methods potentially contain less accurate ordering algorithms. A combination of the ordering methods from the top linear methods, combined with the topology inference from top non-linear methods, could therefore be a possibility for future research.

<!-- Topological complexity versus performance -->
Despite their similar overall performance, the topologies predicted by the top methods differed considerably. Trajectories detected by slingshot and cellTree tended to contain less branches, while those detected by TSCAN and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`). Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data (`r ref("fig", "topology_complexity_comparison", "a")`). Trajectories generated by Monocle DDRTree tended to contain more nodes and edges (`r ref("fig", "topology_complexity_comparison", "b")`), which could give this method an advantage on datasets with complex tree topologies, but could also explain its relatively low performance on linear and bifurcating datasets. Indeed, when we assessed how often a method inferred the correct topology, slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree was by far the best method to infer tree topologies. Nonetheless, the overall accuracy of topology prediction was very low, with slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in 12% of the cases. Inferring the correct topology without parameter tuning is therefore still an open challenge. Conversely, when the data contains a complex trajectory structure, TI will currently still require a considerable guidance by the user, to either optimise the parameters or to choose the method whose output best fits the user's expert knowledge.

@@

`r ref("fig", "topology_complexity_comparison")` **Comparing the ability of the top TI methods to detect the correct trajectory type.** a) % of datasets on which a TI method detects a particular trajectory type, compared with the correct trajectory type. b) Distributions of the topological complexity, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data.

<!-- Additional factors which could influence performance -->
Apart from the trajectory types, we also investigated several other factors which could influence the performance of a method. Although we provided prior information to those methods which require it, we did not observe any major difference in method performance between these methods and methods that did not receive any prior information (`r ref("fig", "results_overview", "b")`). When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as slingshot and SCORPIUS), k-means (which include slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance ( `r ref("stable", "component_importance")`). On the other hand, methods using t-SNE and ICA for their dimensionality reduction were ranked lower, although this was not statistically significant.

@@

## Practical guidelines

Based on the results of our benchmark, we created a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`). As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will by primarily driven by the prior knowledge of the user about what trajectory topology is expected in the data. For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple trajectories, cycles or a complex tree structure. In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree performing best when the data contained a complex tree, and slingshot performing equally well on less complex trajectories . No methods dealing with multiple trajectories or cycles were included in the current version of the evaluation, although two recent methods, Waddington-OT`r cite_methods("wadot")` and AGA`r cite_methods("aga")`. should be able to handle graph trajectories or even disconnected trajectories . In the case where the user would know the exact expected topology, our evaluation suggests the use of reCAT for cycles, SCORPIUS, slingshot and cellTree for linear trajectories, and slingshot, TSCAN and SCOUP for bifurcating trajectories, although certain methods could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`). The most difficult use case is when the topology is known but more complex than a bifurcating or cyclic trajectory. Here, to our knowledge, only ElPiGraph.R ([github.com/Albluca/ElPiGraph.R](https://github.com/Albluca/ElPiGraph.R)), which is not yet published, can be used.

@@

`r ref("fig", "user_guidelines", anchor=TRUE)` **Practical guidelines for method users.** As the performance of a method most heavily depended on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data. We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory. Methods to the right are ranked according to overall performance. Further to the right are shown the user friendly scores (++: ≥ 0.9, +: ≥ 0.8, ± ≥ 0.65, - ≥ 0.5) and overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5).

When choosing a method, it is important to take two further points into account. First, it is important that a trajectory and the downstream results and/or hypotheses originating from it are confirmed by multiple TI methods. This to make sure the model is not biased towards the particular model underlying a TI method, for example its preferred trajectory type. Second, even if the expected topology is known, it can be beneficial to also try out methods which make the less assumptions about the trajectory topology. When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the data. When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.

<!-- The context -->
Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. Indeed, for no other single-cell analysis types have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017]. It is therefore critical that these methods, now reaching `r sum(implementations$is_ti, na.rm=TRUE)`, are evaluated to guide users in their choice. In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods. When comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art. We believe that the benchmarking presented in this study will pave the way to the next series improvements in the field of trajectory inference.

>>>>

Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory. Since 2014, at least `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare. As a result, a comprehensive assessment of the performance of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.

In this study, we evaluated a total of  `r nrow(filter(methods, evaluated, type == "algorithm"))` methods in terms of cellular ordering, trajectory topology, code quality and user friendliness. Our results indicate that there is no "one-size-fits-all" method, but that the choice of method mostly depends on the topology present in the data. We therefore developed a set of guidelines to aid users select the most appropriate method, based on the user's prior knowledge of the trajectory present in the data. While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extendable and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).

# Introduction
Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before [@tanay_scaling_2017]. One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014]. Such dynamic processes can be modelled computationally using trajectory inference (TI), which order cells along a trajectory by their overall similarity in -omics datasets (mainly transcriptomics) [@trapnell_definingcelltypes_2015; @cannoodt_computational_2016; @moon_manifoldlearningbasedmethods_2018]. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also allow to identify more complex trajectory topologies such as cyclic or disconnected graphs. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  thereby allowing the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interactions responsible for one or more bifurcations [@aibar_scenicsinglecellregulatory_2017]. 
Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "implementations")`). For no other single-cell analyses workflows have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017]. All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce. Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type of topologies they can detect. Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along this fixed topology. 
Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods to be applicable on more use cases. Methods that perform topology inference are still in the minority, though current trends suggest this will change soon. A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is to our knowledge the only TI method currently able to deal with disconnected graphs.

<!-- §3: Problem statement -->
Given the diversity in TI methods, an important issue to address is a quantitative assessment of their performance and robustness. Many attempts at tackling this issue have already been made [@haghverdi_diffusionpseudotimerobustly_2016; @ji_tscanpseudotimereconstruction_2016; @welch_slicerinferringbranched_2016; @matsumoto_scoupprobabilisticmodel_2016; @duverle_celltreebioconductorpackage_2016; @cannoodt_scorpiusimprovestrajectory_2016; @lonnberg_singlecellrnaseqcomputational_2017; @campbell_probabilisticmodelingbifurcations_2017], but a comprehensive benchmarking evaluation of TI methods, which compares a large set of methods on a large set of datasets, is still lacking. This is problematic, as new users to the field are confronted with a wide array of TI methods, without a clear idea about which method would optimally solve their problem. Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods (`r ref("fig", "evaluation_overview", "a")`). The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("table", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

`r ref("fig", "evaluation_overview")` **Overview of several key aspects of the evaluation.** a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.

@@

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. We then defined a set of metrics for comparing the similarities of such trajectories, each assessing a different aspect of the trajectory: the similarity in cell ordering, the cell neighbourhood and the topology (`r ref("fig", "evaluation_overview", "c")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.

<!-- Overall evaluation results -->
Across all datasets and evaluation metrics, we found that Slingshot, TSCAN, SCORPIUS, AGA pseudotime and cellTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).  When we focused on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS returned the most accurate linear trajectories, Angle and reCAT optimally identified cycles, and Monocle DDRTree, SCOUP and AGA pseudotime most accurately identified trees. This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, TSCAN, SCOUP and AGA, obtaining a better topological similarity score, while other methods, including SCORPIUS, Slingshot, Arctangent and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).

@@

`r ref("fig", "results_overview")` **Overview of the results on the evaluation of `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods and `r nrow(filter(methods, evaluated, type == "control"))` control methods.** a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use. b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation. Three controls (random trajectory, component 1 of a PCA and angle on a PCA) were added, as these simple methods return relatively accurate trajectories, to which other TI methods can be compared c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced. d) Performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation. Also listed are the quality control scores aggregated according to practicality and the different categories.

@@

<!-- Errors-->
Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g. Monocle DDRTree, cellTree Gibbs, SLICE), meaning these methods could rank even higher if not for the failed executions. Methods that do not scale well with respect to the number of cells or genes exceeded the pre-defined time or memory limits for the largest datasets (reCAT, SCOUP, StemID, MFA, cellTree Gibbs). For a few methods, the time or memory limits were exceeded too often, making the benchmarking results difficult to compare to those of other methods (Ouija, Pseudogp).

<!-- Evaluation robustness -->
```{r}
correlation_tenth <- read_rds(result_file("task_sample_correlations.rds", "11-evaluation_robustness")) %>%
filter(n_tasks == round(max(n_tasks)/10)) %>% 
          pull(cor) %>% 
          mean()
top_method_rank_match <- read_rds(result_file("task_sample_ranks_match.rds", "11-evaluation_robustness")) %>%
filter(method_short_name == "slngsht") %>% 
filter(n_tasks == round(max(n_tasks)/10)) %>%
pull(rank_match)
random_method_max_tasks_top <- read_rds(result_file("task_sample_max_tasks_top.rds", "11-evaluation_robustness")) %>% 
filter(method_short_name == "wishbone") %>% 
pull(max_n_tasks_top)
```

We further investigated the robustness of our evaluation across datasets and dataset sources. Method performance was moderately to highly correlated (0.5-0.9 depending on the trajectory type) between real and synthetic datasets (`r ref("sfig", "robustness_comparison", "a")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard. Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`). To further illustrate this, we looked at the scores in a randomly selected 10% subset of all datasets. The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, Slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets. On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best. For example, for Wishbone, a method ranked in the middle, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method. Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large number of datasets, as we included in our evaluation, is necessary to robustly compare the overall performance of methods.

## Relationships between method characteristics and performance

<!-- Inverse relation between edge flip and correlation -->
In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("fig", "results_overview", "b")`). These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type. However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as Embeddr, SCORPIUS and Topslam, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories. This indicates that current non-linear methods potentially contain less accurate ordering algorithms. A combination of the ordering algorithms from the top linear methods, together with the topology inference from top non-linear methods, could therefore be a possibility for future research.

<!-- Topological complexity versus performance -->
Despite their similar overall performance, the topologies predicted by the top methods differed considerably. Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN, AGA Pseudotime and Monocle DDRTree gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`). AGA Pseudotime and Monocle DDRTree for example, frequently predicted a tree topology, even when only a cyclic, linear or bifurcating topology was present in the data. Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology (`r ref("stable", "topology_sensitivity")`). Nonetheless, the overall accuracy of topology prediction was very low, with Slingshot correctly predicting bifurcating and converging topologies for half of the datasets, and Monocle DDRTree predicting the correct tree topology in only 11% of the cases  (`r ref("stable", "topology_sensitivity")`). Inferring the correct topology without parameter tuning is therefore still an open challenge. When the data contains a complex trajectory structure, TI will currently require a considerable guidance by the user, either to optimise the parameters or to choose the method whose output best fits the user's expert knowledge.

@@

`r ref("fig", "topology_complexity_comparison")` **Comparing the ability of the top TI methods to detect the correct trajectory type.** Distributions of the topological complexity, defined by the sum of the number of milestones and edges between milestones, compared with the true trajectory type present in the data.

<!-- Additional factors which could influence performance -->
Apart from the trajectory types, we also investigated several other factors which could influence the performance of a method. Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "b")`). When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance, while those which included tSNE had a lower performance ( `r ref("stable", "component_importance")`).

```{r}
add_stable(
  read_rds(figure_file(
    "component_importance.rds",
    "6-performance_predictors"
  ))[[params$table_format]], 
  "component_importance", 
  "Importance values for algorithmic components. The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a corrected p-value using a two-tailed Mann–Whitney U test and the increase in node purity importance measure using random forest classification."
)
```

@@

<!-- Overview of evaluation -->

<!-- Practical guidelines -->
Based on the results of our benchmark, we created a set of practical guidelines for method users (`r ref("fig", "user_guidelines")`). As a method's performance is heavily dependent on the trajectory type being studied, the choice of method will be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data. For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure. In each of these use cases, a different set of methods performed optimally, with Monocle DDRTree and AGA Pseudotime performing best when the data contained a complex tree, Slingshot and TSCAN performing well on less complex tree-like trajectories, and AGA Pseudotime and AGA performing well when the data can contain a cycle or multiple disconnected trajectories. In the case where the user expects a particular topology, our evaluation suggests the use of Angle (one of our control methods) and reCAT for cycles, SCORPIUS and cellTree for linear trajectories, TSCAN and Slingshot for bifurcating trajectories, and Slingshot and SCOUP for multifurcating trajectories. although certain methods (those with a free topology) could return other topologies if it would fit the data more accurately (`r ref("fig", "user_guidelines")`).

@@

`r ref("fig", "user_guidelines", anchor=TRUE)` **Practical guidelines for method users.** As the performance of a method most heavily depended on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data. We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory. Methods to the right are ranked according to overall performance. Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65) and overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5).

When choosing a method, it is important to take two further points into account. First, it is important that a trajectory and the downstream results and/or hypotheses originating from it are confirmed by multiple TI methods. This to make sure the model is not biased towards the particular model underlying a TI method, for example its preferred trajectory type. Second, even if the expected topology is known, it can be beneficial to also try out methods which make the less assumptions about the trajectory topology. When the expected topology is confirmed using such a method, it provides extra evidence to the user's knowledge of the biological system. When a more complex topology is produced, this could indicate the presence of a more complex trajectory in the data than was expected by the user.

<!-- The robustness of the evaluation -->
While we have shown that the  Large subset of datasets on which a method is best. This raises important issues, even methods in the sub top deserve to be published.

<!-- The context -->
Trajectory inference is unique among most other categories of single-cell analysis methods, such as clustering, normalisation and differential expression, because it models the data in a way that was almost impossible using bulk data. It is therefore critical that these methods, now reaching `r sum(implementations$contains_ti, na.rm=TRUE)`, are evaluated to guide users in their choice. In this study, we presented a large scale evaluation of the performance and code of  `r sum(methods$evaluated)` TI methods. When comparing the maximum overall score over time, it is encouraging to see multiple incremental improvements over the state-of-the-art. We believe that the benchmarking presented in this study will pave the way to the next series improvements in the field of trajectory inference.
